Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass

### 2024-04-23 manuscript.Rmd
add suggestions from alain and marcal

---
title: "Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass"
author: "Camille Piponiot*, Marguerite Cogné*, Vincent Freycon, Alexis Toumazeau, Régis Peltier"
date: "2024-03-19"
output: 
  officedown::rdocx_document:
    reference_docx: word-styles-ref-02.docx
bibliography: references.bib
nocite: |
  @Orsini1976, @Nair1993, @Nair2021
csl: misc/springer-basic-author-date.csl
---

@@

Agroforestry has the potential to make agriculture more resilient while improving carbon sequestration by incorporating trees and other woody perennials into agricultural fields and diversifying agricultural landscapes. Traditional agricultural systems in tropical areas often include trees, but their carbon sequestration potential is not always well described, hindering their potential inclusion in climate change mitigation strategies. In this study, we quantified carbon storage in both vegetation biomass and soil in five traditional agroforestry systems (AFS) in Timor-Leste:  cropping systems with fallow, silvopastures, young agroforests, home gardens and forest gardens. Our results show that these traditional AFS store large amounts of carbon: the AFS with the highest carbon stocks (forest gardens) stored an average of 210\ Mg\ C\ ha$^{-1}$, similar to an old-growth tropical forest. Biomass carbon was strongly dependent on the type of AFS (which differed in tree cover), while soil carbon was less variable between AFS but more dependent on site. We found no relationship between the amount of carbon stored in biomass and soil. Our results highlight the high diversity of traditional agroforestry systems in Timor-Leste and their high capacity to store carbon.

@@

Agricultural expansion has been the main driver of tropical deforestation over the past half century [@Pendrill2022;@Gibbs2010;@Rudel2009], resulting in large greenhouse gas emissions [approximately 2.6 gigatonnes CO~2~e/yr over 2010-2014; @Pendrill2019]. The demand for tropical agricultural land is expected to continue to increase as the population grows [@UN2022]. In addition, tropical areas are expected to be disproportionately affected by climate change, putting their agriculture and food systems at risk [@Lawrence2015]. There is therefore an urgent need to promote tropical agricultural landscapes that are both more resilient to climate change, while maintaining high productivity and ecological value [@Harvey2014]. In particular, increasing the carbon stored in the soil and vegetation of agricultural landscapes could be an effective way to mitigate climate change while increasing the resilience of the food production system [@Lal2004;@Minasny2017].

@@

Agroforestry systems (AFS) have the potential to store large amounts of carbon in the vegetation and soil, and this carbon storage capacity depends on the characteristics of AFS considered [@Albrecht2003;@Feliciano2018]. As trees store large amounts of carbon, the carbon stored in the vegetation of AFS is largely dependent on the number and size of trees in the system [@Ma2020]. However, the effect of AFS on soil organic carbon (SOC) has been relatively little studied [@Beillouin2021], is more variable and depends on a number of factors. Although AFS appear to increase SOC stocks overall, some AFS such as woodlots (i.e. planting trees with intercropping during the establishment phase) may decrease SOC compared to non-AFS, while AFS such as silvopastures have been shown to strongly increase SOC stocks [@Feliciano2018]. Other factors such as climate and previous land use type have also been shown to strongly influence SOC in AFS, which may explain the variability of observations reported in the literature [@Feliciano2018;@Ma2020;@Martin2020]. Furthermore, the relationship between the amount of carbon stored in aboveground vegetation and soil in AFS could only be demonstrated for the topsoil layer [< 10 cm; @Ma2020].

In Timor-Leste, a large proportion of the population depends on agriculture for their livelihoods, but food security is threatened by population growth, increasing pressure on natural resources, and an increasingly variable climate with an unreliable rainfall regime [@WB2009;@Molyneux2012]. In this context, productive, resilient and sustainable agriculture is critical for the country, and agroforestry has been presented as a potential solution to many of the challenges facing Timorese agriculture [@Paudel2022;@Cogne2024]. Although trees already form a large part of traditional agricultural systems in Timor-Leste [@Paudel2022;@Cogne2024] and can be key elements for biodiversity conservation in the tropics [@Perfecto2008;@Torquebiau1992], they have received little attention and, to our knowledge, no study has quantified carbon storage in traditional Timorese AFS. Better quantification of these carbon stocks would improve the understanding of the ecological importance of these AFS at both local and global scales, and help to integrate them into national and international climate change mitigation initiatives.

In this study, we aim to better understand traditional agroforestry systems in Timor-Leste, and in particular their carbon storage both in their biomass and in the soil. We hypothesize that different types of traditional agroforestry systems will exhibit measurable differences in aboveground carbon (AGC; H1a) and SOC (H1b) stocks. We also assume, following @Ma2020, that there is a positive relationship between AGC and SOC stocks in the topsoil layer (0-10 cm; H2a), but no relationship with SOC below 10 cm (H2b).

Agroforestry has the potential to make agriculture more resilient while improving carbon sequestration by incorporating trees and other woody perennials into agricultural fields and diversifying agricultural landscapes. Traditional agricultural systems in tropical areas often include trees, but their carbon sequestration potential is not always well described, hindering their potential inclusion in climate change mitigation strategies. In this study, we quantified carbon storage in both vegetation biomass and soil in five traditional agroforestry systems (AFS) in Timor-Leste:  cropping systems with fallow, silvopastures, young agroforests, home gardens and forest gardens. Our results show that these traditional AFS store large amounts of carbon: the AFS with the highest carbon stocks (forest gardens) stored an average of 255 MgC/ha, similar to an old-growth tropical forest. Biomass carbon was strongly dependent on the type of AFS (which differed in tree cover), while soil carbon was less variable between AFS but more dependent on site. We found no relationship between the amount of carbon stored in biomass and soil. Our results highlight the high diversity of traditional agroforestry systems in Timor-Leste and their high capacity to store carbon.

@@

# simplify site names
data_plot[, site := substr(ID_VILLAGE, 1, 1)]
data_plot[site == "S", site := "Osso Luga"]
data_plot[site == "G", site := "Gariuai"]
```

@@

The research was carried out in the center of the Baucau district, on the north-eastern coast of Timor-Leste. Two communities, or suco (group of villages), were selected to represent two contrasting topo-geographical, climatic and historical cases: (i) Gariuai suco (long-lat coordinates: `r site_summary[site=="Gariuai", coord]`) is located on a plateau of Baucau limestone, with an annual rainfall of `r site_summary[site=="Gariuai", prec]` mm and an altitude ranging 300-750 m; and (ii) the village of Osso Luga in Samalari suco (long-lat coordinates: `r site_summary[site=="Osso Luga", coord]`) is located on the Matebian foothill based on Bobonaro Scaly Clay, with an annual rainfall of `r site_summary[site=="Osso Luga", prec]` mm and an altitude ranging `r site_summary[site=="Osso Luga", elev]` m (Figure\ \@ref(fig:map-sites)). The soils of the Baucau district are classified using the WRB soil classification [@Mantel2023] as Lithosols and Calcic Luvisols [@FAO2012]. The average soil properties differed between villages (Table\ S\@ref(tab:site-soil)). Gariuai had a clay loam texture whereas Osso Luga had a clay texture. The two villages also have had different histories. Gariuai has been inhabited for at least 50 years, while Osso Luga was abandoned during the Indonesian occupation, and was only repopulated after 2003. During the period of abandonment, the various agroforestry systems in Osso Luga lay fallow.

@@

We used the typology described in @Cogne2024, developed in the same study area. This typology was based on the one proposed by Nair et al. (1993, 2021) and adapted to the local context with semi-structured interviews conducted in the same study area [@Cogne2024]. The five AFSs identified were: Crop system including a wooded Fallow phase (CF), Silvopasture (SP), Young Agroforest (YA), Home Garden (HG) and Forest Garden (FG). CFs are plantations of a few non-perennial crops (such as maize, sweet potatoes, groundnuts and squash) which are left fallow after 1 to 3 years. SPs consist in a vast area (generally 200 to 500 ha) where cattle are grazed under a layer of trees. YAs are agricultural plots where the farmer has been planting rows of trees, spaced at around 5x10m, for two to 10 years, usually with the help of projects. HGs are plots of land close to houses, where crops and livestock (poultry, pigs, cattle, etc.) have been grown in the shade of numerous trees and palms for multiple uses (wood, fruit, fibre, etc.) for more than five years. FGs are generally former HGs (over 50 years old), which have been abandoned, for example as a result of the war, and are now used mainly for gathering with limited human influence.

The AFSs are presented in more detail in the supplemental information (Table\ S\@ref(tab:afs-typology)) and in @Cogne2024.

@@

We randomly selected 3 plots from each of the 5 traditional AFSs identified in each of the two sites (30 plots in total; Figure\ \@ref(fig:map-sites)). Sampling plots were defined as the entire agricultural plot (area between 0.06-1.25 ha) for all AFSs except SP plots, which were generally much larger than 1 ha. In SP plots, a 1 ha sampling plot was delineated within the agricultural plot. Tree and soil measurements (described below) were carried out between June and August 2021.

@@

Taxonomic identification was carried out using tetum vernacular names, which were then linked to scientific names using the iNaturalist and Pl\@ntNet applications, Geoffrey Hull’s monograph [@Hull2006], the “Useful tropical plants” database [@Fern2014] as well as the Australian Northern Territory government's Native Plants Herbarium [@NT2013]. The correspondence between tetum names and scientific names is provided in the supplemental material (Table S\@ref(tab:tetum-names)).

@@

In each of the 30 sample plots, three composite samples were taken from three of the five 10 x 10 m subplots, following a transect covering the within plot variability and at a minimum distance of 1 m from any tree (Figure\ S\@ref(fig:design)). Soils were sampled with an auger at 0-10 cm and 20-30 cm. Soils samples were then air-dried and sieved at 2 mm. They were subjected to standard physical and chemical analyses at the Cirad laboratories in Montpellier, France (Table\ S\@ref(tab:site-soil)). Among these analyses, total organic carbon was determined by dry combustion and measured with a thermal conductive detector.

In addition, at one location at the center of the plot, soils were sampled with a cylinder of 250 cm$^3$ at 2-7 cm, 12-17 cm and 22-27 cm. Each sample was dried at 105°C in a classic oven until reaching a stable dry mass, sieved to 2 mm, and then the fine fraction (< 2 mm) was weighed.

@@

We estimated tree-level AGC stocks from the measured tree CBH and height using the pantropical allometric equation from @Chave2014 and the R package BIOMASS [@Rejou-Mechain2017]. Each tree was assigned a mean and standard deviation of wood density at the species or genus level when at least one value was available from the Global Wood Density Database [@Zanne2009]. Other taxa were assigned a plot-level average wood density. The biomass of the palms was estimated by multiplying their estimated volume by their estimated wood density. The volume of the palm trunks was estimated as the volume of a cylinder, using the height and diameter measured in situ, with a volumetric shrinkage of 11% [@WD2007]. The height of the palms was measured above the leaves and was therefore multiplied by a conversion factor of 0.9 to obtain the trunk height. Biomass was then converted into carbon with a factor of 0.4713 [@Thomas2012].

Tree-level belowground carbon (BGC) stocks were estimated using the root-to-shoot ratio equation from @Ledo2018, which is based on tree DBH and climatic water deficit from @Chave2014. This ratio was then multiplied by the estimated aboveground carbon stock to obtain tree-by-tree belowground carbon stock.

@@

$$SOC_i = \frac{\sum_i Corg_i \cdot (mfine_i \cdot 10^{-6}) \cdot (Splot \cdot \Delta d)}{Vcyl}$$

@@

All statistical analyses were carried out using R [@R2023].

@@

The estimated AGC and SOC stocks were log-transformed for all statistical tests, because we expect them to be log-normally distributed as they only take positive values. We removed one CF plot at Gariuai from the following statistical tests, as it did not contain any trees exceeding the minimum measurement size (CBH = 30 cm) and its AGC stocks were therefore estimated at 0\ Mg\ C\ ha$^{-1}$ (even if there were trees < 30 cm CBH) and could not be log-transformed. After testing for normality (Shapiro-Wilks test) and homoscedasticity (Levene’s test), we performed a two-way ANOVA on the log-transformed estimates of AGC and SOC stocks, grouped by AFS and site. To test the relationship between log-transformed SOC and AGC stocks, we fitted two linear mixed models. We predicted SOC stocks at two different depths (either 0-10 or 0-30 cm) as a function of AGC stocks (fixed effect) and site (Samalari or Osso-Luga; random effect).

@@

The uncertainty on plot-level AGC was calculated with the AGBmonteCarlo() function from the BIOMASS package, using a (rather conservative) coefficient of variation of 15% on height measurements [@Hunter2013]; large and small errors on 5 and 95% of trees respectively [@Chave2004;@Rejou-Mechain2017].
To propagate uncertainty on SOC estimates, we used a coefficient of variation of 12% for organic carbon (Cirad laboratory in Montpellier, France), 10% for the mass of the fine fraction [@Page1999].

@@

Total carbon stocks varied between `r dataC_plot[variable=="TOT", round(min(mean))]` and `r dataC_plot[variable=="TOT", round(max(mean))]`\ Mg\ C\ ha$^{-1}$, with an average of `r dataC_plot[variable=="TOT", round(mean(mean))]`\ Mg\ C\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="TOT", round(sd(mean)/mean(mean)*100)]`%  (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)). The AFS with the lowest total carbon stocks was the `r dataC_afs[variable=="TOT", afs[which.min(mean)]]` with an average of `r dataC_afs[variable=="TOT", round(mean[which.min(mean)])]`\ Mg\ C\ ha$^{-1}$; the AFS with the highest total carbon stocks was the `r dataC_afs[variable=="TOT", afs[which.max(mean)]]` with an average of `r dataC_afs[variable=="TOT", round(mean[which.max(mean)])]`\ Mg\ C\ ha$^{-1}$ (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)).

@@

AGC stocks varied between `r dataC_plot[variable=="AGC", round(min(mean))]` and `r dataC_plot[variable=="AGC", round(max(mean))]`\ Mg\ C\ ha$^{-1}$, with an average of `r dataC_plot[variable=="AGC", round(mean(mean))]`\ Mg\ C\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="AGC", round(sd(mean)/mean(mean)*100)]`%  (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)). AGC stocks varied significantly between AFS (ANOVA p-value: `r subset(anova_table, variable=="AGC" & factor == "afs")$pvalue`, Table\ \@ref(tab:anova)), but not between site (ANOVA p-value = `r subset(anova_table, variable=="AGC" & factor == "site")$pvalue`, Table\ \@ref(tab:anova)). The AFS with the lowest AGC stocks was the `r dataC_afs[variable=="AGC", afs[which.min(mean)]]` with an average of `r dataC_afs[variable=="AGC", round(mean[which.min(mean)])]`\ Mg\ C\ ha$^{-1}$; the AFS with the highest AGC stocks was the `r dataC_afs[variable=="AGC", afs[which.max(mean)]]` with an average of `r dataC_afs[variable=="AGC", round(mean[which.max(mean)])]`\ Mg\ C\ ha$^{-1}$ (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)).

@@

SOC stocks varied between `r dataC_plot[variable=="SOC", round(min(mean))]` and `r dataC_plot[variable=="SOC", round(max(mean))]`\ Mg\ C\ ha$^{-1}$, with an average of `r dataC_plot[variable=="SOC", round(mean(mean))]`\ Mg\ C\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="SOC", round(sd(mean)/mean(mean)*100)]`%  (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)). No significant differences in SOC were observed between the five AFS studied (ANOVA p-value: `r subset(anova_table, variable=="SOC" & factor == "afs")$pvalue`, Table\ \@ref(tab:anova)). However, the measured SOC stocks differed between sites (ANOVA p-value: `r subset(anova_table, variable=="SOC" & factor == "site")$pvalue`, Table\ \@ref(tab:anova)). The SOC stocks in Gariuai were on average higher (`r dataC_site[site=="Gariuai" & variable=="SOC", round(mean)]`\ Mg\ C\ ha$^{-1}$) than those in Osso Luga (`r dataC_site[site=="Osso Luga" & variable=="SOC", round(mean)]`\ Mg\ C\ ha$^{-1}$; Figure\ \@ref(fig:barplot)).

@@

The relationship between AGC and SOC was not significant (p-value = `r round(dfreg$pvalue[grepl("30", dfreg$depth)], 2)`; Figure\ \@ref(fig:plot-lmereg)). This relationship remained insignificant (p-value = `r round(dfreg$pvalue[grepl("10", dfreg$depth)], 2)`) when tested with SOC stocks in the topsoil layer (0-10 cm). The proportion of variance explained by the linear mixed models were `r round(dfreg$R2_cond[grepl("30", dfreg$depth)]*100)`% and `r round(dfreg$R2_cond[grepl("10", dfreg$depth)]*100)`% for 0-30 and 0-10 cm depths respectively, but a marginal R$^2$ of just `r round(dfreg$R2_marg[grepl("30", dfreg$depth)]*100, 1)`% and `r round(dfreg$R2_marg[grepl("10", dfreg$depth)]*100, 1)`% respectively when only the fixed effect was considered (Figure\ \@ref(fig:plot-lmereg)). Overall, AGC stocks were a poor predictor of SOC stocks.

@@

The AFS studied here vary significantly in terms of AGC stocks, in agreement with our hypothesis H1a. This reflects the density and size of trees in the plot, which in turn depends on the importance of trees in the production system. In some systems, such as CF and SP, trees do not contribute to food production but have other functions, such as restoring soil fertility (e.g. during the fallow period) or providing shade for livestock. This more limited role for trees may explain why their total biomass is lower in these systems: in the CF system, the plot is regularly cleared and annual crops are planted in full sun, while in the SP system, the growth of grazing grasses is favored by reducing the number of trees that shade them.

>>>>

---
title: "Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass"
author: "Camille Piponiot*, Marguerite Cogné*, Vincent Freycon, Alexis Thoumazeau, Marçal Gusmão, Régis Peltier"
date: "2024-03-19"
output: 
  officedown::rdocx_document:
    base_format: "bookdown::word_document2"
    reference_docx: word-styles-ref-02.docx
bibliography: references.bib
nocite: |
  @Orsini1976, @Nair1993, @Nair2021
csl: misc/springer-basic-author-date.csl
---

@@

Agroforestry has the potential to make agriculture more resilient while improving carbon sequestration by incorporating trees and other woody perennials into agricultural land and diversifying landscapes. Traditional agricultural systems in tropical areas often include trees, but their carbon sequestration potential is not always well described, hindering their inclusion in climate change mitigation strategies. In this study, we quantified carbon storage in both vegetation biomass and soil in five traditional agroforestry systems (AFSs) in Timor-Leste, namely: cropping systems with fallow, silvopastures, young agroforests, home gardens and forest gardens. Our results show that these traditional AFSs can store large amounts of carbon: they stored an average of 156\ Mg\ C\ ha$^{-1}$, and the AFSs with the highest carbon stocks (forest gardens) stored an average of 210\ Mg\ C\ ha$^{-1}$, close to the values of old-growth tropical forests. Biomass carbon was strongly dependent on the type of AFS (which differed in tree cover), while soil carbon was less variable between AFS but more dependent on site. We found no relationship between the amount of carbon stored in biomass and soil. Our results highlight the high diversity of traditional AFSs in Timor-Leste and their high carbon sequestration capacity. These results could provide an important baseline for the inclusion of AFS in Timor-Leste's climate change mitigation strategy, and could serve as a reference for future AFS studies in different agro-climates of Timor-Leste.

@@

Agricultural expansion has been the main driver of tropical deforestation over the past half century [@Pendrill2022;@Gibbs2010;@Rudel2009], resulting in large greenhouse gas emissions [approximately 2.6 gigatonnes CO~2~e/yr over 2010-2014; @Pendrill2019]. The demand for tropical agricultural land is expected to continue to grow as population increases [@UN2022]. In addition, tropical areas are expected to be disproportionately affected by climate change, putting their agriculture and food systems at risk [@Lawrence2015]. There is therefore an urgent need to promote tropical agricultural landscapes that are both more resilient to climate change, while maintaining high productivity and ecological value [@Harvey2014]. In particular, increasing the amount of carbon stored in the soil and vegetation of agricultural landscapes could be an effective way to mitigate climate change while increasing the resilience of the food production system [@Lal2004;@Minasny2017].

@@

Agroforestry systems (AFSs) have the potential to store large amounts of carbon in the vegetation and soil, and this carbon storage capacity depends on the characteristics of AFS considered [@Albrecht2003;@Feliciano2018]. As trees can store large amounts of carbon, the carbon stored in the vegetation of AFSs is largely dependent on the number and size of trees in the system [@Ma2020]. However, the effect of AFSs on soil organic carbon (SOC) has been relatively little studied [@Beillouin2021], is more variable and depends on a number of factors. Although AFSs appear to increase SOC stocks overall, some AFS such as woodlots (i.e. planting trees with intercropping during the establishment phase) may decrease SOC stocks compared to non-AFSs, while AFSs such as silvopastures have been shown to strongly increase SOC stocks [@Feliciano2018]. Other factors such as climate and previous land use type can also influence SOC in AFSs, which may explain the variability of observations reported in the literature [@Feliciano2018;@Ma2020;@Martin2020]. Furthermore, the relationship between the amount of carbon stored in aboveground vegetation and soil in AFS could only be evidenced for the topsoil layer [< 10 cm; @Ma2020].

In Timor-Leste, a large proportion of the population depends on agriculture for their livelihoods, and food security is threatened by population growth, increasing pressure on natural resources, and an increasingly variable climate with an unreliable rainfall regime [@WB2009;@Molyneux2012]. In this context, productive, resilient and sustainable agriculture is critical for the country, and agroforestry has been presented as a potential solution to many of the challenges facing Timorese agriculture [@Paudel2022;@Cogne2024]. Although trees already form a large part of traditional agricultural systems in Timor-Leste [@Paudel2022;@Cogne2024] and can be key elements for biodiversity conservation [@Perfecto2008;@Torquebiau1992], they have received little attention and, to our knowledge, no study has quantified carbon storage in traditional Timorese AFSs. A precise quantification of these carbon stocks would improve the understanding of the ecological importance of these AFSs at both local and global scales, and help to integrate them into national and international climate change mitigation initiatives.

In this study, we aim to better understand traditional AFSs in Timor-Leste by quantifying the carbon storage in both biomass and soil of five traditional AFSs: cropping systems with fallow, silvopastures, young agroforests, home gardens and forest gardens. We hypothesize that different types of traditional AFSs will exhibit measurable differences in aboveground carbon (AGC; H1a) and SOC (H1b) stocks. We anticipate, following @Ma2020, that there is a positive relationship between AGC and SOC stocks in the topsoil layer (0-10 cm; H2a), but no relationship with SOC below 10 cm (H2b).

@@

# simplify site names
data_plot[, site := substr(ID_VILLAGE, 1, 1)]
data_plot[site == "S", site := "Osso-Luga"]
data_plot[site == "G", site := "Gariuai"]
```

@@

The research was carried out in the center of the Baucau municipality, on the north-eastern coast of Timor-Leste. Two communities, or suco (group of villages), were selected to represent two contrasting topo-geographical, climatic and historical cases: (i) Gariuai suco (long-lat coordinates: `r site_summary[site=="Gariuai", coord]`) located on a plateau of Baucau limestone, with an annual rainfall of `r site_summary[site=="Gariuai", prec]` mm and an altitude ranging 300-750 m above sea level (asl); and (ii) the village of Osso-Luga in Samalari suco (long-lat coordinates: `r site_summary[site=="Osso-Luga", coord]`) located on the Matebian foothill based on Bobonaro Scaly Clay, with an annual rainfall of `r site_summary[site=="Osso-Luga", prec]` mm and an altitude ranging `r site_summary[site=="Osso-Luga", elev]` m asl (Figure\ \@ref(fig:map-sites)). The soils of the Baucau municipality are classified using the WRB soil classification [@Mantel2023] as Lithosols and Calcic Luvisols [@FAO2012]. The average soil properties differed between villages (Table\ S\@ref(tab:site-soil)): Gariuai had a clay loam texture whereas Osso-Luga had a clay texture. The two villages also have different recent histories: Gariuai has been inhabited for at least 50 years, while Osso-Luga was abandoned during the Indonesian occupation and was only repopulated after Timor-Leste gained independence in 2002. During the period of abandonment, the AFSs in Osso-Luga lay fallow.

@@

We used the typology described in @Cogne2024, which was developed in the same study area. This typology was based on the one proposed by Nair et al. (1993, 2021) and adapted to the local context with semi-structured interviews conducted in the same study area [@Cogne2024]. The five AFSs identified were: Crop system including a wooded Fallow phase (CF), Silvopasture (SP), Young Agroforest (YA), Home Garden (HG) and Forest Garden (FG). CFs are plantations of a few non-perennial crops (such as maize, sweet potatoes, groundnuts and squash) which are left fallow after 1 to 3 years. SPs consist in a vast area (generally 200 to 500 ha) where cattle are grazed under a layer of trees. YAs are agricultural plots where the farmer has been planting rows of trees, spaced at around 5x10m, for two to 10 years. HGs are plots of land located close to houses, where crops and livestock (poultry, pigs, cattle, etc.) have been grown in the shade of numerous trees and palms for multiple uses (wood, fruit, fibre, etc.) for more than five years. FGs are generally former HGs (over 50 years old), which have been abandoned, for example as a result of the war, and are now used mainly for gathering with limited human influence.

The AFSs are presented in more detail as supplemental information (Table\ S\@ref(tab:afs-typology)) and in @Cogne2024.

@@

We randomly selected 3 plots from each of the 5 traditional AFSs identified in each of the two sites (30 plots in total; Figure\ \@ref(fig:map-sites)). Sampling plots were defined as the entire agricultural plot (area between 0.06 and 1.25 ha) for all AFSs except SP plots, which were generally much larger than 1 ha. In SP plots, a 1 ha sampling plot was delineated within the agricultural plot. Tree and soil measurements (described below) were carried out between June and August 2021.

@@

Taxonomic identification was carried out using Tetum  (a national official language) vernacular names, which were then linked to scientific names using the iNaturalist and Pl\@ntNet applications, Geoffrey Hull’s monograph [@Hull2006], the “Useful tropical plants” database [@Fern2014] as well as the Australian Northern Territory government's Native Plants Herbarium [@NT2013]. The correspondence between Tetum names and scientific names is provided in the supplemental material (Table S\@ref(tab:tetum-names)).

@@

In each of the 30 sample plots, three composite samples were taken from three of the five 10 x 10 m subplots, following a transect covering the intra-plot variability and at a minimum distance of 1 m from any tree (Figure\ S\@ref(fig:design)). Soils were sampled with a 15cm auger at 0-10 cm and 20-30 cm depth; due to the size of the auger we couldn't accurately sample the intermediate soil layer. Soils samples were then air-dried and sieved at 2 mm. They were subjected to standard physical and chemical analyses at Cirad laboratories in Montpellier, France (Table\ S\@ref(tab:site-soil)). Among these analyses, total organic carbon was determined by dry combustion and measured with a thermal conductive detector.

In addition, at one location at the center of the plot, soils were sampled with a cylinder of 250 cm$^3$ at 2-7 cm, 12-17 cm and 22-27 cm. Each sample was dried at 105°C in a classic oven until reaching a stable dry mass, sieved to 2 mm, and then the fine fraction (< 2 mm) was weighed to be used in SOC stocks calculation (equation \@ref(eq:soc)).

@@

We estimated tree-level AGC stocks from the measured tree CBH and height using the pantropical allometric equation from @Chave2014 and the R package BIOMASS [@Rejou-Mechain2017]. Each tree was assigned a mean and standard deviation of wood density at the species or genus level when at least one value was available from the Global Wood Density Database [@Zanne2009]. Other taxa were assigned a plot-level average wood density. The biomass of the palms was estimated by multiplying their estimated volume by their estimated wood density. The volume of the palm trunks was estimated as a cylinder, using the height and diameter measured *in situ*, with a volumetric shrinkage of 11% [@WD2007]. The height of the palms was measured above the leaves and was therefore multiplied by a conversion factor of 0.9 to obtain the trunk height. Biomass was then converted into carbon with a factor of 0.4713 [@Thomas2012].

Belowground carbon (BGC) stocks were estimated at tree level using the root-to-shoot ratio equation from @Ledo2018, which is based on tree DBH and climatic water deficit from @Chave2014. This ratio was then multiplied by the estimated AGC stock to obtain tree-by-tree BGC stock.

@@

\begin{equation}
SOC_i = \frac{\sum_i Corg_i \cdot (mfine_i \cdot 10^{-6}) \cdot (Splot \cdot \Delta d)}{Vcyl} (\#eq:soc)
\end{equation}

@@

All statistical analyses were carried out using the R software [@R2023].

@@

Estimated AGC and SOC stocks were log-transformed for all statistical tests, because we expected them to be log-normally distributed as they only take positive values. We removed one CF plot at Gariuai from the following statistical tests, as it did not contain any trees exceeding the minimum measurement size (CBH = 30 cm) and its AGC stocks were therefore estimated at 0\ Mg\ C\ ha$^{-1}$ (even if there were trees < 30 cm CBH) and could not be log-transformed. After testing for normality (Shapiro-Wilks test) and homoscedasticity (Levene’s test), we performed a two-way ANOVA on the log-transformed estimates of AGC and SOC stocks, grouped by AFS and site. To test the relationship between log-transformed SOC and AGC stocks, we fitted two linear mixed models. We predicted SOC stocks at two different depths (either 0-10 or 0-30 cm) as a function of AGC stocks (fixed effect) and site (Samalari or Osso-Luga; random effect).

@@

The uncertainty on plot-level AGC was calculated with the AGBmonteCarlo function from the R package BIOMASS, using a (rather conservative) coefficient of variation of 15% on height measurements [@Hunter2013]; large and small errors on 5 and 95% of trees respectively [@Chave2004;@Rejou-Mechain2017].
To propagate uncertainty on SOC estimates, we used a coefficient of variation of 12% for organic carbon (as recommended by Cirad Laboratories in Montpellier, France), 10% for the mass of the fine fraction [@Page1999].

@@

Total carbon stocks ranged between `r dataC_plot[variable=="TOT", round(min(mean))]` and `r dataC_plot[variable=="TOT", round(max(mean))]`\ Mg\ C\ ha$^{-1}$, with an average of `r dataC_plot[variable=="TOT", round(mean(mean))]`\ Mg\ C\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="TOT", round(sd(mean)/mean(mean)*100)]`%  (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)). The AFS with the lowest total carbon stocks was the `r dataC_afs[variable=="TOT", afs[which.min(mean)]]` with an average of `r dataC_afs[variable=="TOT", round(mean[which.min(mean)])]`\ Mg\ C\ ha$^{-1}$; the AFS with the highest total carbon stocks was the `r dataC_afs[variable=="TOT", afs[which.max(mean)]]` with an average of `r dataC_afs[variable=="TOT", round(mean[which.max(mean)])]`\ Mg\ C\ ha$^{-1}$ (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)).

@@

AGC stocks ranged between `r dataC_plot[variable=="AGC", round(min(mean))]` and `r dataC_plot[variable=="AGC", round(max(mean))]`\ Mg\ C\ ha$^{-1}$, with an average of `r dataC_plot[variable=="AGC", round(mean(mean))]`\ Mg\ C\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="AGC", round(sd(mean)/mean(mean)*100)]`%  (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)). AGC stocks varied significantly between AFS (ANOVA p-value: `r subset(anova_table, variable=="AGC" & factor == "afs")$pvalue`, Table\ \@ref(tab:anova)), but not between site (ANOVA p-value = `r subset(anova_table, variable=="AGC" & factor == "site")$pvalue`, Table\ \@ref(tab:anova)). The AFS with the lowest AGC stocks was the `r dataC_afs[variable=="AGC", afs[which.min(mean)]]` with an average of `r dataC_afs[variable=="AGC", round(mean[which.min(mean)])]`\ Mg\ C\ ha$^{-1}$; the AFS with the highest AGC stocks was the `r dataC_afs[variable=="AGC", afs[which.max(mean)]]` with an average of `r dataC_afs[variable=="AGC", round(mean[which.max(mean)])]`\ Mg\ C\ ha$^{-1}$ (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)).

@@

SOC stocks varied between `r dataC_plot[variable=="SOC", round(min(mean))]` and `r dataC_plot[variable=="SOC", round(max(mean))]`\ Mg\ C\ ha$^{-1}$, with an average of `r dataC_plot[variable=="SOC", round(mean(mean))]`\ Mg\ C\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="SOC", round(sd(mean)/mean(mean)*100)]`%  (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)). No significant differences in SOC were observed between the five AFS studied (ANOVA p-value: `r subset(anova_table, variable=="SOC" & factor == "afs")$pvalue`, Table\ \@ref(tab:anova)). However, the measured SOC stocks differed between sites (ANOVA p-value: `r subset(anova_table, variable=="SOC" & factor == "site")$pvalue`, Table\ \@ref(tab:anova)). The SOC stocks in Gariuai were on average higher (`r dataC_site[site=="Gariuai" & variable=="SOC", round(mean)]`\ Mg\ C\ ha$^{-1}$) than those in Osso-Luga (`r dataC_site[site=="Osso-Luga" & variable=="SOC", round(mean)]`\ Mg\ C\ ha$^{-1}$; Figure\ \@ref(fig:barplot)).

@@

The relationship between AGC and SOC was not significant (p-value = `r round(dfreg$pvalue[grepl("30", dfreg$depth)], 2)`; Figure\ \@ref(fig:plot-lmereg)). This relationship remained insignificant (p-value = `r round(dfreg$pvalue[grepl("10", dfreg$depth)], 2)`) when tested with SOC stocks in the topsoil layer (0-10 cm). The proportion of variance explained by the linear mixed models were `r round(dfreg$R2_cond[grepl("30", dfreg$depth)]*100)`% and `r round(dfreg$R2_cond[grepl("10", dfreg$depth)]*100)`% for 0-30 and 0-10 cm depths respectively, but a marginal R$^2$ of just `r round(dfreg$R2_marg[grepl("30", dfreg$depth)]*100, 1)`% and `r round(dfreg$R2_marg[grepl("10", dfreg$depth)]*100, 1)`% respectively when only the fixed effect was considered (Figure\ \@ref(fig:plot-lmereg)).

@@

The studied AFSs were found to differ significantly in terms of AGC stocks, in agreement with our hypothesis H1a. This result reflects the density and size of trees in the plot, which in turn depends on the importance of trees in the production system. In some systems, such as CF and SP, trees do not contribute directly to food production but have other functions, such as restoring soil fertility (e.g. during the fallow period) or providing shade for livestock. This more limited role for trees may explain why their total biomass is lower in these systems: in the CF system, the plot is periodically cleared and the remaining trees are pruned to allow the planted annual crops to receive full sun, while in the SP system, the growth of grazing grasses is favored by reducing the number of trees.
